### 2018-05-27 4_impl.tex
Cleanup finale up to Xevent.

The main call for commencing the work on \porthos[2] was the need for processing real-world C programs, which, at first, requires the input language to be extended.
This implies the support not only for new syntactic structures of the C language (such as the \texttt{switch} statement or the postfix increment operator \texttt{i++}), but also for its fundamental concepts and features (such as types, pointer arithmetic or first-order functions), which requires revision of the whole architecture of the tool.
Yet far from all the C language is supported (which, considering its complexity and numerous pitfalls, goes far beyond current thesis%
%
\footnote{To ensure this, one merely has to look at existing C compiler implementations, for instance, the open-source \tool{gcc} compiler, which uses the C parser written in more than 18.5 thousand lines of code (see \url{https://github.com/gcc-mirror/gcc/blob/master/gcc/c/c-parser.c})}%
%
), we consider the accomplished work as a step towards it.
%This applies not only to instances that simply constitute the syntactic sugar of C language (such as the \texttt{switch} statement or the prefix increment operator \texttt{++i}), but also to its fundamental concepts and features (such as pointer arithmetic or structures).
%However, the architecture of current version of \porthos{} does not allow to ... as it is ....
% Although and optimisation of the tool so that it is able to process real C programs in perspective.

%In current Chapter we discuss the new architecture of \porthos[2].

% function invocations/calls: inlining with binding
% contexts: for now, no ctxs
% comparing to cprover (https://www.cprover.org/cbmc/doc/manual.pdf, page 45, absatz about 'goto') where "the loop is unwound a given number of times" , we SHOULD have 2 modes of unrolling: 1) number of times, and 2) number of high-level instructions executed
% TODO(code): (https://www.cprover.org/cbmc/doc/manual.pdf, page 45) "The break and continue statements are replaced by equivalent goto statements as described in the ANSI-C standard."

% TODO: arrays: cprover manual page 48

%\section{The architecture of \porthos{}}
\section{General principles}
\label{ch:impl:principles}

The previous version of \porthos{} did not distinguish the event-based program model from the high-level AST, they both were encoded into a single SMT-formula (see classes of package `\texttt{dartagnan.program}' of \porthos[1]).
%For 
Moreover, the syntax tree was implemented as a mutable data structure, which is being modified at all stages of the program (for instance, see the methods `\texttt{dartagnan.program.Program.compile(...)}' of \porthos{} that recursively compute some properties of the AST and change its state).
We are inclined to consider this architecture as one that is fast to develop, but hard to maintain (since it is difficult to guarantee the correctness of the program) and extend (since adding the support for a new high-level instruction requires changing multiple components of the program, from parser to encoder).
%Moreover, we encounter serious problems when we need to add support for control-flow jumps (as \texttt{continue}, \texttt{break}, \texttt{goto} in C).

Therefore, while working on the new design of \porthos[2], we decided to clearly separate the high-level intermediate code representation (an AST structure) from the low-level event-based representation (an event-flow graph).
Such a modular architecture will allow to support multiple input languages%
%
\footnote{Apart from the C language, \porthos[2] must be able to analyse litmus tests in different assembly languages and, as a compatibility mode, the input language of \porthos[1].} %
%
by parsing them and converting parsed syntax trees to a simplified AST.

All internal representations used by \porthos[2] must be immutable, so that it is possible to guarantee the correctness of the program by controlling its invariants.
The immutability in \porthos[2] is implemented via \texttt{final} fields that are assigned by the immutable-object values (either a primitive type, or another immutable object, or an immutable collection provided by the library Guava by Google%
%
\footnote{The Guava project repository: \url{https://github.com/google/guava/}}%
%
).

@@

\vspace{0.5em}
\begin{enumerate}[nolistsep]
  \item \textit{Robustness:}
    \begin{enumerate}[label*=\arabic*.]
      \item preservation the completeness of analysis,
      \item modular architecture: each module can be tested independently,
      \item usage of software design patterns where necessary, and
      \item usage of immutable data structures for all DTOs.
    \end{enumerate}
  \item \textit{Transparency:}
    \begin{enumerate}[label*=\arabic*.]
      \item following the principles of simplicity and readability,
      \item clear and informative program output, and
      \item following the clear code style.
    \end{enumerate}
  \item \textit{Efficiency:}
    \begin{enumerate}[label*=\arabic*.]%[leftmargin=1.5em]
      \item keeping the trade-off between execution time and memory usage.
    \end{enumerate}
  \item \textit{Extensibility:}
    \begin{enumerate}[label*=\arabic*.]%[leftmargin=1em]
      \item clear modular architecture.
    \end{enumerate}
\end{enumerate}

Robustness of the analysis is the main criterion of \porthos[2] as a verification tool.
Although it makes the analysis more sensitive to the combinatorial explosion, it preserves the completeness of analysis, which is necessary for a model-checking tool.
As a robust and transparent tool, \porthos[2] must adhere to the strategy of aborting its work on any unexpected outcome (for instance, if a parser failed to parse the string and the recovery algorithm is not described).
One of the transparency principles is following the clear code style.
This mainly means the clear and informative naming of classes, functions and fields.
It also implies the unified ordering of methods in classes, minimised size of methods code, clear tabulation, etc.
%The modular architecture of \porthos[2] allow each its module to be tested independently.
%Usage of immutable data structures for all DTOs helps 
%immut: preservation of invariants
%The transparency, efficiency and extensibility are the necessary criteria of any good tool.

As its predecessor, \porthos[2] uses the open-source SMT solver \tool{Z3}%
%
\footnote{The \tool{Z3} project repository: \url{https://github.com/Z3Prover/z3}} %
%
by Microsoft Research~\cite{de2008z3}. However, unlike its predecessor, \porthos[2] has an additional abstraction level, \zformula{} (see Section~\ref{ch:impl:model:zformula}), that allows the use any other SMT solver.

%We have decided to implement \porthos[2] in \textit{Java}
As its predecessor, \porthos[2] is written in \textit{Java},
%The programming language choice for \porthos[2] was also made in favour of \textit{java},
firstly, in order to be able to reuse some parts and concepts of \porthos{} also written in Java, and secondly, because the authors find the concepts of Object-Oriented Programming~(OOP) in Java suitable for modelling languages.
Although Java does not show the best results in performance benchmarks (for example, compared to C++~\cite{hundt2011loop}), the performance cornerstone of \porthos[2] (as well as any other SMT-based code analyser) is the phase of solving the SMT-formula, which is left to the third-party SMT solver invoked by \porthos[2] via a Java API.
However, considering the perspective of using \porthos[2] as a static analyser for real-world programs, the memory optimisation problem must also be taken into account during both encoding and solving stages.
It is worth noting that, for the reasons of simplicity, \porthos[2] is not a concurrent program, however, we believe that, due to its modular architecture, it can be easily parallelised on the level of program modules.

Currently, \porthos[2] can operate only in the intra-procedural analysis mode, assuming that each function defined in the input file is being executed in a separate thread.
%TODO: describe more intra- and inter-procedural!
However, the redesigned architecture of \porthos[2] can be easily extended to support the inter-procedural (cross-procedure) analysis that inlines function calls and binds variable contexts.
%TODO : check in the implementation , THAT WE HAVE NOT IMPLEMENTED THIS FEATURE!!
Thus, instead of analysing a single source code file, the user can analyse the whole code project.
In this mode, the functions to be executed in parallel should be specified by the user.
Also, the tool may be extended to detect the concurrent parts of the code automatically at the pre-compilation stage.
For that, the pre-compiler should recognise functions that commence a new process (such as \texttt{pthread\_create} from \texttt{pthread.h}) and resolve the argument that points to the thread function.
This functionality %implies an additional stage of the pre-compilation and 
is left beyond current thesis.
%resolve the semantics of function invocations and remember the functions 
%function that commence new process or thread 
%(for instance, the function \texttt{pthread\_create} of the library \texttt{pthread.h} that starts the new thread of the function it received as the first argument).
%TODO: inter: user specifies the repository with code, functions that intended to work in parallel and runs.

%TODO: say sth about 100500 litmus-tests for kernel

\section{Program input}
\label{ch:impl:input}
%(for instance, the original input language of \porthos{} tool, two variants of syntax of a litmus test used by \tool{herd}, an assembly language for any supported architecture).

Both \porthos{} and \porthos[2] use the \tool{ANTLR} parser generator%
%
\footnote{The \tool{ANTLR} project repository: \url{https://github.com/antlr/antlr4}}%
%
~\cite{parr2013definitive}, a powerful language processing tool.
The \tool{ANTLR} takes as input the user-defined grammar of the target language in a BNF-like form and produces the LL(*)-parser and optionally some auxiliary classes (such as listeners and visitors for the syntax tree).
Although this parser may not be as efficient as a hand-written language-optimised parser, it reduces the overhead of implementing the parser significantly.
Among other advantages \tool{ANTLR}, it has a rather large collection of officially supported grammars.
Nonetheless, the intuitive syntax for defining grammars and numerous of tools for debugging grammars make the \tool{ANTLR} an attractive instrument for solving the parsing problem.

\begin{figure}[t]%[!h]%[!hb]
\centering
%basicstyle=\ttfamily\tiny,
\begin{lstlisting}[basicstyle=\rmfamily\scriptsize,%
  %alsoletter={:,|,;,(,)},
  %morekeywords={:,|,;,(,)},
  %keywordstyle=\color{dkblue},
  literate=%
  {<}{{$\langle$}}1
  {>}{{$\rangle$}}1,
  escapeinside={[*}{*]},]
<program>
      :      <initialisation>   <thread>+   <assertion>
      ;
<thread>   
      :      thread   <thread-id>   <instruction>
      ;
<instruction>
      :      <atom>
      |      `{'   <instruction>   `}'
      |      <instruction>   `;'   <instruction>
      |      `while'   `('   <bool-expr>   `)'   <instruction>
      |      `if'   <bool-expr>   `{'   <instruction>   `}'   <instruction>
      ;
<atom>   
      :      <register>   `[*<*]-'   <expression>
      |      <register>   `[*<*]:-'   <location>
      |      <location>   `:='   <register>
      |      <register>   `='   <location>   `.'   `load'      `('   <atomic>   `)'
      |      <location>   `='   <register>   `.'   `store'   `('   <atomic>   `)'
      |      (`mfence'   |   `sync'   |   `lwsync'   |   `isync')
      ;
<bool-expr>
      :      `true'
      |      `false'
      |      <expression>   (`and'      |      `or')   <expression>
      |      <expression>   (`=='   |   `!='   |   `>'   |   `>='   |   `[*<*]'   |   `[*>*]=')   <expression>
      ;
<expression>   
      :      [0-9]
      |      <register>
      |      <expression>   (`*'   |   `+'   |   `-'   |   `/'   |   `%')   <expression>
      ;
\end{lstlisting}
\caption{The sketch of the input language grammar used by \oldporthos}
\label{fig:in_grammar_pts}
\end{figure}

Figure~\ref{fig:in_grammar_pts} represents the grammar sketch in BNF syntax of the input language used by \porthos[1].
%(the full \tool{ANTLR} grammar is available at Appendix~\ref{apx:grammar}).
%TODO! Reformulate this.
The input language parser used by \porthos{} suffered from several disadvantages.
Firstly, it contained the parser code inlined directly into the grammar, so that the grammar would serve as a template for the parser code (which is called semantic actions).
Such a combining of two expressive languages makes the code hardly understandable and, therefore, poorly maintainable.
In \porthos[2], we clearly separated the parser (generated from the grammar file `\textit{<grammar>.g4}') from converting the \tool{ANTLR} syntax tree to the AST, that is one for all languages of an input program.

Secondly, \porthos{} resolved the semantics of operations syntactically (it was defined in the \tool{ANTLR} grammar), whereas it should be resolved by a separate module operating on the AST level%
%
%\footnote{Most modern languages (including C++) allow the function overloading, thus requiring the type information to be available for the function resolution algorithm (see Section~\ref{ch:impl:proc:x-pre-compiler}).}%
%
, so that it does not require to change grammar for encoding the semantics of a new function. %adding support for a new function.
%\footnote{for instance, given two functions `\lstinline{foo(int a)}' and `\lstinline{foo(char a)}' compiled by a C++ compiler, the code `\lstinline{int a = '1'; foo(a);}' will invoke the first method rather than the second one.}.%
As the reader may have noticed from the grammar sketch in Figure~\ref{fig:in_grammar_pts}, the memory operations of different kinds vary syntactically as well.
For example, the assignment of local computation to a register uses the symbol `\lstinline{<-}', the atomic non-relaxed load operation denoted as `\lstinline{<:-}', atomic non-relaxed store operation denoted as `\lstinline{:=}', and the semantics of relaxed \lstinline{load} and \lstinline{store} are resolved syntactically by matching the function name.
Moreover, only the operator `\lstinline{<-}' could have an expression as the source of data, which means that expressions could be assigned only to registers.
% (thus, the variable \lstinline{r} was recognised as a local variable as well).
In \porthos[2], the semantics of the data-flow operation is determined according to the types of operands, that are determined during the pre-compilation stage (see Section~\ref{ch:impl:proc:x-pre-compiler}).
The semantics of the functions also is being resolved during the pre-compilation stage via the \textit{invocation hooking} mechanism (see Section~\ref{ch:impl:proc:x-compiler}).

Thirdly, the grammar used by \porthos{} allowed only a restricted set of operations.
For example, it accepted the computation expressions only over local variables.
Thus, in the assignment expression `\lstinline{r <- (x + 1);}', the variable \lstinline{x} was parsed as a local variable even though it can be used as a shared variable in other parts of the program.
In \porthos[2], all shared variables involved into a computation expression are tentatively copied to temporary local variables.
%Moreover, expressions could be assigned only to local variables only via the assignment operator `\lstinline{<-}' (thus, the variable \lstinline{r} was recognised as a local variable as well).
%, which might lead to inconsistency of the result SMT-formula if the same variable name was used both as a register and as a location, for instance, the in code `\lstinline{x := reg1; reg2 <- (x + 1);}', the first statement interprets the variable \lstinline{x} as a location, while the grammar of second statement requires it to be a register.

Fourthly, \porthos[1] supported only integer constants and expressions.
%Also, only integers were processed by the input language parser of \porthos{}.
In \porthos[2], we extended support for primitive types supported by the \tool{Z3} solver (this apply to 32-bit integers encoded as \texttt{Int}s of \tool{Z3}, floats encoded as \texttt{Real}s, enumerations encoded as \texttt{Scalar}s).
%TODO : constant arrays? https://rise4fun.com/\tool{Z3}/tutorial/guide
Although the \tool{Z3} supports the array theory (characterised by the select-store axioms~\cite{de2011z3}), the complexity of pointer analysis for the arrays of non-constant size moves the full support of arrays and pointers out of the scope of current thesis.
%Thus, adding support for more advanced structures as pointers, arrays, function definitions and calls was one of the purposes of revising the tool architecture.

Finally, the grammar used by \porthos{} had the following minor drawbacks. %A minor drawback of 
The operators were implemented as non-associative (expressions of the form `\lstinline{1 + 2 * 3}' could not be parsed).
Comparing to C statements, that must end with the semicolon punctuator `\lstinline{;}', statements defined in the grammar of \porthos[1] use the semicolon as a separator between statements (the final statement must not end with semicolon).
The litmus-specific syntax for variables initialisation was used only for declaring the shared variables (all of them were initialised with default value \lstinline{0}), however, this syntax should be used as an initial assignment of both shared and local variables with arbitrary values.
%initialise them with a value different from the default \lstinline{0} or to initialise a local variable.
%did not allow to assign an arbitrary value to a variable (the default value \lstinline{0} was used).

%Also, \porthos{} supports only the litmus-specific syntax for the variables initialisation, however, it allows only initialisation of the shared variable and only by default value \lstinline{0}.
%\porthos[2] supports arbitrary declaration to be performed in initialisation statement.

@@

The general architecture scheme of \porthos[2] is presented in Figure~\ref{fig:arch}.
%The rectangles in the picture denote processing units (marked with gear sign with a unique number of the component), and labels of the arrows describe the internal representation carried out during the transformation.

%TODO: ATTENTION HERE TO THE PICTURE!!!!!!!!!!!!!!!!!!!!!!!!
\begin{figure}[H]
  \centering
  \includegraphics[height=.95\textheight,keepaspectratio]{img/my/draw.io/general_arch-no_numbering.png}
  \caption{The general architecture of \porthos[2]}
  \label{fig:arch}
\end{figure}

The program takes as input the program to be analysed and one (the reachability analysis mode) or two (the portability analysis mode) memory models.
However, instead of parsing memory models from the \cat{} file, the tool may operate with pre-defined memory models (\texttt{SC}, \texttt{TSO}, \texttt{PSO}, \texttt{RMO}, \texttt{ALPHA}, \texttt{POWER}, \texttt{ARM}; this feature is inherited from \porthos[1]).
The parsed program syntax tree is then converted (Section~\ref{ch:impl:proc:y-constr}) to a program AST called \ytree{}%
%
\footnote{In order to avoid confusion between different internal representations, we prefix the names of elements of each internal representation with a letter. For instance, we picked the letter `Y' to denote the AST code representation as drawing of this letter resembles the tree branching; with letter `X' we prefix elements of the event-flow graph as the events are to be e\textbf{x}ecuted; and with letter `W' we prefix elements of the \textbf{w}eak memory model AST.}%
%
(Section~\ref{ch:impl:model:ytree}), which then is being preprocessed at the pre-compilation stage (Section~\ref{ch:impl:proc:x-pre-compiler}) in order to collect information necessary for the compilation.
The \ytree{} then is being compiled (Section~\ref{ch:impl:proc:x-compiler}) to an \xgraph{} representation (Section~\ref{ch:impl:model:xgraph}).
The compiled \xgraph{} then is being converted to an acyclic form (Section~\ref{ch:impl:proc:x-unroll}) in order to be encoded into a \zformula{} (Section~\ref{ch:impl:model:zformula}).
Apart from that, the memory-model constructor (Section~\ref{ch:impl:proc:w-constr}) constructs the abstract syntax tree of derived relations of the weak memory model \wmodel{} (Section~\ref{ch:impl:model:wmodel}).
%on the basis of the \xgraph{} in order to build the weak memory model \wmodel{} (Section~\ref{ch:impl:model:wmodel}).
Thereafter, \wmodel{} and acyclic \xgraph{} are encoded (Section~\ref{ch:impl:proc:z-encoder}) to a \zformula{} representation (a wrapper over an SMT-formula), which then is translated to an SMT-formula (Section~\ref{ch:impl:proc:smt-converter}), which then is solved by the SMT solver (Section~\ref{ch:impl:proc:smt-solver}).

@@

%Internal representations used in \porthos[1]

%As it has already been mentioned, all internal representations used by \porthos[2] are immutable.
For keeping the architecture transparent, we build all abstraction levels with interfaces, even if some of them does not add any new functionality.% (as java does not support multiple inheritance for classes (only for interfaces), we

@@

The first internal representation used by \porthos[2] is the \textit{\ytree{}}, which represents a high-level recursively%
%
\footnote{Hereinafter we use the term \textit{recursive} data structure (sometimes called \textit{inductive}) to refer a complex data type which can contain elements that contain other elements of the same type.
For example, an unary expression contains references to the operator and the child expression of the same type.} %
%
defined AST.
The Apendix~\ref{apx:trees} presents the file tree of the main classes that constitute the \ytree{} hierarchy (as the inheritance tree might be obvious for the C-like AST, we confine ourselves to presenting the classes file tree only, which we tend to retain clearly structured).

>>>>

The main call for commencing the work on \porthos[2] was the need for processing real-world C programs, which, at first, requires the input language to be extended.
This implies the support not only for new syntactic structures of the C language (such as the \texttt{switch} statement or the postfix increment operator \texttt{i++}), but also for its fundamental concepts and features (such as types, pointer arithmetic or first-order functions), which requires revision of the whole architecture of the tool.
Yet far from all the C language is supported (which, considering its complexity and numerous pitfalls, goes far beyond current thesis%
%
\footnote{To ensure this, one merely has to look at existing C compiler implementations, for instance, the open-source \tool{gcc} compiler, which uses the C parser written in more than 18.5 thousand lines of code (see \url{https://github.com/gcc-mirror/gcc/blob/master/gcc/c/c-parser.c})}%
%
), we consider the accomplished work as a step towards it.

\section{General principles}
\label{ch:impl:principles}

The \porthos[1] does not distinguish the event-based program model from the high-level AST, and they both are encoded into a single SMT-formula (see classes of package `\texttt{dartagnan.program}' of \porthos[1]).
Moreover, the syntax tree is implemented in \porthos[1] as a mutable data structure, which is being modified at all stages of the program (for instance, see the methods `\texttt{dartagnan.program.Program.compile(...)}' of \porthos[1] that recursively compute some properties of the AST and change its state).
We are inclined to consider the old architecture to be fast to develop, but difficult to maintain (since it is difficult to guarantee the correctness of the program) and extend (since adding the support for a new high-level instruction requires changing multiple components of the program, from parser to encoder).

Therefore, while working on the new design of \porthos[2], we clearly separated the high-level intermediate code representation (the AST structure) from the low-level event-based representation (the event-flow graph).
Such a modular architecture will allow to support multiple input languages%
%
\footnote{Apart from the C language, \porthos[2] should be able to analyse litmus tests in different assembly languages and, as a compatibility mode, the input language of \porthos[1].} %
%
by parsing them and converting parsed syntax trees to a simplified AST.

All data-transfer objects~(DTOs) used by \porthos[2] must be immutable, so that it is possible to guarantee the correctness of the program by controlling preservation of its invariants.
The immutability in \porthos[2] is implemented via \texttt{final} fields that are assigned by the immutable-object values (either a primitive type, or another user-defined immutable object, or an immutable collection provided by the library Guava by Google%
%
\footnote{The Guava project repository: \url{https://github.com/google/guava/}}%
%
).

@@

\vspace{0.5em}
\begin{enumerate}[nolistsep]
  \item \textit{Robustness:}
    \begin{enumerate}[label*=\arabic*.]
      \item completeness of the analysis,
      \item modular architecture: each module can be tested independently,
      \item usage of software design patterns where necessary, and
      \item usage of immutable data structures for all DTOs.
    \end{enumerate}
  \item \textit{Transparency:}
    \begin{enumerate}[label*=\arabic*.]
      \item following the principles of simplicity and readability,
      \item clear and informative program output, and
      \item following the clear code style.
    \end{enumerate}
  \item \textit{Efficiency:}
    \begin{enumerate}[label*=\arabic*.]%[leftmargin=1.5em]
      \item keeping the trade-off between execution time and memory usage.
    \end{enumerate}
  \item \textit{Extensibility:}
    \begin{enumerate}[label*=\arabic*.]%[leftmargin=1em]
      \item clear modular architecture.
    \end{enumerate}
\end{enumerate}

Robustness of the analysis is the main criterion of \porthos[2] as a verification tool.
Although it makes the analysis more sensitive to the combinatorial explosion, it preserves the completeness of analysis, which is necessary for a model-checking tool.
As a robust and transparent tool, \porthos[2] must adhere to the strategy of aborting its work on any unexpected outcome (for instance, if a parser failed to parse the string and the recovery algorithm is not described).
One of the transparency principles is following the clear code style.
This mainly means the clear and informative naming of classes, functions and fields.
It also implies the unified ordering of methods in classes, minimised size of methods code, clear tabulation, etc.

As its predecessor, \porthos[2] uses the open-source SMT-solver \tool{Z3}%
%
\footnote{The \tool{Z3} project repository: \url{https://github.com/Z3Prover/z3}} %
%
by Microsoft Research~\cite{de2008z3}. However, unlike its predecessor, \porthos[2] has an additional abstraction level, \zformula{} (see Section~\ref{ch:impl:model:zformula}), that allows the use any other SMT-solver.

As its predecessor, \porthos[2] is written in \textit{Java}, firstly, in order to be able to reuse some parts and concepts of \porthos{} also written in Java, and secondly, because the authors find the concepts of Object-Oriented Programming~(OOP) inherent in Java suitable for modelling the programming languages.
Although Java does not show the best results in performance benchmarks (for example, compared to C++~\cite{hundt2011loop}), the performance cornerstone of \porthos[2] (as well as any other SMT-based code analyser) is the phase of solving the SMT-formula, which is left to the third-party SMT-solver invoked by \porthos[2] via a Java API.
However, considering the perspective of using \porthos[2] as a static analyser for real-world programs, the memory optimisation problem must also be taken into account during both encoding and solving stages.
It is worth noting that, for the reasons of simplicity, \porthos[2] is not a concurrent program, however, we believe that due to its modular architecture it can be easily parallelised on the level of program modules.

Currently, \porthos[2] can operate only in the intra-procedural analysis mode (not supporting invocations of the user-defined functions), assuming that each function defined in the input file is being executed in a separate thread.
However, the redesigned architecture of \porthos[2] can be easily extended to support the inter-procedural (cross-function) analysis by inlining user-defined functions calls and binding variable contexts.
In this mode, instead of analysing a single source code file, the tool can process the whole code project, where the functions to be executed in parallel are specified separately by the user.
Also, the tool can be extended to detect the concurrent parts of the code automatically at the pre-compilation stage.
For that, the pre-compiler should recognise functions that commence a new process (such as \texttt{pthread\_create} from \texttt{pthread.h}) and resolve the argument that points to the thread function.
This functionality is left beyond current thesis.

\section{Program input}
\label{ch:impl:input}

Both \porthos[1] and \porthos[2] use the \tool{ANTLR} parser generator%
%
\footnote{The \tool{ANTLR} project repository: \url{https://github.com/antlr/antlr4}}%
%
~\cite{parr2013definitive} for creating the input-language parsers.
The \tool{ANTLR} takes as input the user-defined grammar of the target language in a BNF-like form and produces the LL(*)-parser and optionally some auxiliary classes (such as listeners and visitors for the syntax tree).
Although this parser may not be as efficient as a hand-written language-optimised parser, it reduces the overhead of implementing the parser significantly.
Among other advantages \tool{ANTLR}, it has a rather large collection of officially supported grammars.
Nonetheless, the intuitive syntax for defining grammars and numerous of tools for debugging grammars make the \tool{ANTLR} an attractive instrument for processing languages.

\begin{figure}[t]%[!h]%[!hb]
\centering
%basicstyle=\ttfamily\tiny,
\begin{lstlisting}[basicstyle=\rmfamily\scriptsize,%
  %alsoletter={:,|,;,(,)},
  %morekeywords={:,|,;,(,)},
  %keywordstyle=\color{dkblue},
  literate=%
  {<}{{$\langle$}}1
  {>}{{$\rangle$}}1,
  escapeinside={[*}{*]},]
<program>
      :      <initialisation>   <thread>+   <assertion>
      ;
<thread>
      :      thread   <thread-id>   <instruction>
      ;
<instruction>
      :      <atom>
      |      `{'   <instruction>   `}'
      |      <instruction>   `;'   <instruction>
      |      `while'   `('   <bool-expr>   `)'   <instruction>
      |      `if'   <bool-expr>   `{'   <instruction>   `}'   <instruction>
      ;
<atom>   
      :      <register>   `[*<*]-'   <expression>
      |      <register>   `[*<*]:-'   <location>
      |      <location>   `:='   <register>
      |      <register>   `='   <location>   `.'   `load'    `('   <atomic>   `)'
      |      <location>   `='   <register>   `.'   `store'   `('   <atomic>   `)'
      |      (`mfence'   |   `sync'   |   `lwsync'   |   `isync')
      ;
<bool-expr>
      :      `true'
      |      `false'
      |      <expression>   (`and'   |   `or')   <expression>
      |      <expression>   (`=='   |   `!='   |   `>'   |   `>='   |   `[*<*]'   |   `[*>*]=')   <expression>
      ;
<expression>
      :      [0-9]
      |      <register>
      |      <expression>   (`*'   |   `+'   |   `-'   |   `/'   |   `%')   <expression>
      ;
\end{lstlisting}
\caption{Sketch of the input language grammar used by \oldporthos{}}
\label{fig:in_grammar_pts}
\end{figure}

Figure~\ref{fig:in_grammar_pts} represents the grammar sketch in BNF syntax of the input language used by \porthos[1].
The input language parser used by \oldporthos{} suffers from several disadvantages.
Firstly, it contains the parser code inlined directly into the grammar, so that the grammar serves as a template for the parser code (which is called semantic actions in \tool{ANTLR}).
Such a combining of two expressive languages makes the code hardly understandable and, therefore, poorly maintainable.
In \porthos[2], we clearly separated the parser (generated from the grammar file `\textit{<grammar>.g4}') from converting the \tool{ANTLR} syntax tree to the AST, that is one for all languages of an input program.

Secondly, \porthos{} resolves the semantics of operations syntactically (it was defined in the \tool{ANTLR} grammar), whereas it should be resolved by a separate module operating on the AST level%
%
%TODO: check or remove this
%\footnote{Most modern languages (including C++) allow the function overloading, thus requiring the type information to be available for the function resolution algorithm (see Section~\ref{ch:impl:proc:x-pre-compiler}).}%
%
, so that it does not require to change grammar for encoding the semantics of a new function. %adding support for a new function.
%\footnote{for instance, given two functions `\lstinline{foo(int a)}' and `\lstinline{foo(char a)}' compiled by a C++ compiler, the code `\lstinline{int a = '1'; foo(a);}' will invoke the first method rather than the second one.}.%
As the reader might have noticed from the grammar sketch in Figure~\ref{fig:in_grammar_pts}, different kinds of memory operations of of the \oldporthos{} input language vary syntactically as well.
For example, the assignment of a local computation to a register uses the symbol `\lstinline{<-}', the atomic non-relaxed load operation is denoted by \lstinline{<:-}', atomic non-relaxed store operation is denoted by `\lstinline{:=}', and the semantics of relaxed \lstinline{load} and \lstinline{store} is resolved syntactically by matching the function name and arguments.
Moreover, only the operator `\lstinline{<-}' accepts an expression as the source of data, which means that expressions could be assigned only to registers.
In \porthos[2], the semantics of the data-flow operation is determined according to the types of operands, that are determined during the pre-compilation stage (see Section~\ref{ch:impl:proc:x-pre-compiler}).
The semantics of the functions also being resolved during the compilation stage via the \textit{invocation hooking} mechanism (see Section~\ref{ch:impl:proc:x-compiler}).

Thirdly, the grammar used by \oldporthos{} accepts only a restricted set of operations.
For example, the computation expressions can use only local variables.
Thus, in the assignment expression `\lstinline{r <- (x + 1);}', the variable \lstinline{x} will be parsed as a local variable even though it can be used as a shared variable in other parts of the program, which may lead to the inconsistency of analysis.
In \porthos[2], all shared variables involved into a computation expression are tentatively copied to temporary local variables.

Fourthly, \porthos[1] supports only integer constants and expressions.
In \porthos[2], we extended support for primitive types supported by the \tool{Z3} solver (32-bit integers are encoded as \texttt{Int}s of \tool{Z3}, floats are encoded as \texttt{Real}s).
Although the \tool{Z3} supports the array theory (characterised by the select-store axioms~\cite{de2011z3}), the complexity of pointer analysis for the arrays of non-constant size moves the full support of arrays and pointers out of the scope of current thesis.

Finally, the grammar used by \porthos{} has the following minor drawbacks. %A minor drawback of 
The operators are non-associative (expressions of the form `\lstinline{1 + 2 * 3}' can not be parsed).
Comparing to C statements, which must end with the semicolon punctuator `\lstinline{;}', statements defined in the grammar of \porthos[1] use the semicolon as a separator between statements (the final statement must not end with semicolon).
The litmus test-specific syntax for variables initialisation is used only for declaring the shared variables (all of them are initialised with default value \lstinline{0}), however, this syntax should be used as an initial assignment of both shared and local variables with arbitrary values.

@@

\begin{figure}%[H]
  \centering
  \includegraphics[height=.98\textheight,keepaspectratio]{img/my/draw.io/general_arch-no_numbering.png}
  \caption{The general architecture of \porthos[2]}
  \label{fig:arch}
\end{figure}

The general processing scheme of old tool \porthos[1] is the following.
The input program is parsed to the AST, where leaves are the events.
The instructions of the AST are then encoded into the $\phi_{CF}$ (see Section~\ref{ch:enc:bmc:cf}) and the events are processed by the single function (`\texttt{dartagnan.wmm.Domain.encode}') that loops over all events, checks their properties (the kind of event, the operands of the memory event, the type of the barrier event, etc.) and adds corresponding relations assertions to the formula.
Encoding all the relations (the \textit{program domain encoding}) in a single function is error-prone and hardly manageable.
Moreover, this function does not consider the memory model, therefore the result formula can contain relations that are not constrained by the memory model specified by the user and thus are redundant (for instance, the \texttt{TSO} memory model does not contain \texttt{POWER}-specific relations \texttt{ii}, \texttt{ic}, \texttt{ci}, etc., that are always encoded regardless the memory model).
In \porthos[2], we moved the encoding logic into a separate module \ref{ch:impl:proc:z-encoder} that accepts the input program and the user-defined memory model, therefore does not add constraints over unused relations to the result SMT-formula.

The high-level architecture of \porthos[2] is presented in Figure~\ref{fig:arch}.
The program takes as input the program to be analysed and one (the reachability analysis mode) or two (the portability analysis mode) memory models.
However, instead of parsing memory models from the \cat{} file, the tool may operate with pre-defined memory models (\texttt{SC}, \texttt{TSO}, \texttt{PSO}, \texttt{RMO}, \texttt{ALPHA}, \texttt{POWER}, \texttt{ARM}; this feature is inherited from \porthos[1]).
The parsed program syntax tree is then converted (Section~\ref{ch:impl:proc:y-constr}) to a program AST called \ytree{}%
%
\footnote{In order to avoid confusion between different internal representations, we prefix the names of elements of each internal representation with a letter. For instance, we picked the letter `Y' to denote the AST code representation as drawing of this letter resembles the tree branching; with letter `X' we prefix elements of the event-flow graph as the events are to be e\textbf{x}ecuted; and with letter `W' we prefix elements of the \textbf{w}eak memory model AST.}%
%
(Section~\ref{ch:impl:model:ytree}), which then is being preprocessed at the pre-compilation stage (Section~\ref{ch:impl:proc:x-pre-compiler}) in order to collect information necessary for the compilation.
The \ytree{} then is being compiled (Section~\ref{ch:impl:proc:x-compiler}) to an \xgraph{} representation (Section~\ref{ch:impl:model:xgraph}).
The compiled \xgraph{} then is being converted to an acyclic form (Section~\ref{ch:impl:proc:x-unroll}) in order to be encoded into a \zformula{} (Section~\ref{ch:impl:model:zformula}).
Apart from that, the memory-model constructor (Section~\ref{ch:impl:proc:w-constr}) constructs the abstract syntax tree of derived relations of the weak memory model \wmodel{} (Section~\ref{ch:impl:model:wmodel}).
%on the basis of the \xgraph{} in order to build the weak memory model \wmodel{} (Section~\ref{ch:impl:model:wmodel}).
Thereafter, \wmodel{} and acyclic \xgraph{} are encoded (Section~\ref{ch:impl:proc:z-encoder}) to a \zformula{} representation (a wrapper over an SMT-formula), which then is translated to an SMT-formula (Section~\ref{ch:impl:proc:smt-converter}), which then is solved by the SMT-solver (Section~\ref{ch:impl:proc:smt-solver}).

@@

For keeping the architecture transparent, we build all abstraction levels with interfaces, even if some of them does not add any new functionality.

@@

The first internal representation used by \porthos[2] is the \textit{\ytree{}}, which represents an untyped high-level recursively defined AST%
%
\footnote{Hereinafter we use the term \textit{recursive} data structure (also called \textit{inductive} data structure) to refer a complex data type that can contain elements of the same type.} %
%
. The Apendix~\ref{apx:trees} presents the file tree of the main classes that constitute the \ytree{} hierarchy (as the inheritance tree might be obvious for the C-like AST, we confine ourselves to presenting the classes file tree only, which we tend to retain clearly structured).
